A White Noise Approach to Evolutionary Ecology

### 2020-04-02 .A White Noise Approach To Evolutionary Ecology/draft.Rmd
easy_fig3

``\int_0^t\int_\mathbb{R}f(x,s)\dot W(x,s)dxds"=``\int_\mathbb{R}\int_0^tf(x,s)\dot W(x,s)dxds"=\mathbf W_t(f), \ \forall \ f\in\mathscr{N}_2, \ t\geq0.

@@

To obtain a SPDE from a BBM we take a diffusion limit. There are several ways to do this, but a simple approach is to rescale the mass of individuals and time by $1/n$, diffusion by $\mu\to\mu/n$, branching rate by $\lambda\to n\lambda$, fitness by $\mathscr{W}\to\mathscr{W}^{1/n}$ and consider the limit as $n\to\infty$. Denoting the rescaled process by $X_t^{(n)}(D)$, the limiting process $\mathscr{X}_t=\lim_{n\to\infty}X_t^{(n)}$ is called a super-Brownian motion and is also a non-negative finite measure-valued process [@Watanabe1968]. However, instead of returning the number of individuals alive in a region of space, super-Brownian motion returns the _mass_ of the population concentrated in a region of space. Since we have rescaled individual mass by $1/n$ and took the limit $n\to\infty$, individuals are no longer discrete units. The particle view of the population gets replaced by a continuous blob spread across $d$-dimensional space. In particular, the value of $\mathscr{X}_t(D)$ is a continuously varying non-negative random variable for any $t\geq0$ and $D\subset\mathbb{R}^d$. It turns out that for spatial dimension $d=1$, $\mathscr{X}_t$ is absolutely continuous with respect to the Lebesgue measure for each $t\geq0$ [@Konno1988; @Reimers1989]. This means that we can write $\mathscr{X}_t(D)=\int_D \nu(x,t)dx$ for some density process $\nu(x,t)$. Setting $\lambda=1$ and $m=\ln\mathscr{W}$ this density process satisfies the SPDE
\begin{equation}\label{neutral_SPDE}
\frac{\partial}{\partial t}\nu(x,t)=m\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)+\sqrt{V\nu(x,t)}\dot W(x,t).
\end{equation}
Since $\nu(x,t)$ is not generally differentiable in $x$ or $t$, the derivatives in expression (\ref{neutral_SPDE}) are taken in the _weak_ sense. That is, to rigorously interpret SPDE (\ref{neutral_SPDE}), we integrate the solution $\nu(x,t)$ against functions $f\in C_b^2(\mathbb{R})$, where $C_b^2(\mathbb{R})$ is the set of bounded and twice continuously differentiable functions on $\mathbb{R}$. Hence, equation (\ref{neutral_SPDE}) is just an abbreviation for

\begin{multline}\label{neutral_int}
\int_\mathbb{R}\nu(x,t)f(x)dx-\int_\mathbb{R}\nu(x,0)f(x)dx=\int_0^t\int_\mathbb{R}\nu(x,s)\left(mf(x)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}f(x)\right)dsdx \\
+\int_0^t\int_\mathbb{R}f(x)\sqrt{V\nu(x,s)}\dot W(x,s)dxds, \ \forall \ f\in C_b^2(\mathbb{R}).
\end{multline}
This expression is referred to as the _mild_ solution of (\ref{neutral_SPDE}). Evans [-@lawrenceevans2010] provides a nice introduction to weak derivatives. For more on the general theory of SPDE see Walsh [-@Walsh] and Da Prato and Zabczyk [-@DaPrato2014]. Note that since $\nu(x,t)$ is the density of a finite measure, it is integrable for each $t\geq0$. Thus, since for some $M>0$, $|f(x)|\leq M$ for every $x\in\mathbb{R}$, setting $\varphi(x,t)=f(x)\sqrt{V\nu(x,t)}$ implies $\varphi\in\mathscr{N}_2$. Hence, the white noise integral on the right-hand side of equation (\ref{neutral_int}) can be understood using the heuristics introduced above. Evaluating equation (\ref{neutral_int}) in the particular case of $f(x)\equiv1$ returns the total mass process, which we refer to as the total abundance $N(t)$.

@@

where
\begin{equation}
\bar m(t)=\frac{1}{N(t)}\int_\mathbb{R}m(\nu,x)\nu(x,t)dx,
\end{equation}
and
\begin{equation}
\int_0^td\hat{\mathbf{W}}_s(\sqrt{\nu(x,s)})=\int_0^t\int_\mathbb{R}\frac{\sqrt{\nu(x,s)}}{\sqrt{\int_\mathbb{R}\nu(x,s)dx}}\dot W(x,s)dxds.
\end{equation}
Setting $W_1(t)=\hat{\mathbf{W}}_t(\sqrt{\nu(x,t)})$, we can use traditional stochastic differential notation to write
\begin{equation}
dN=\bar mNdt+\sqrt{VN}dW_1.
\end{equation}
To find the associated SDE for $\bar x(t)$ and $\sigma^2(t)$, we want to repeat the same approach for $f(x)=x,x^2$ and apply It&ocirc;'s lemma. However, for these cases $f\notin C^2_b(\mathbb{R})$ since $f$ will not be bounded. Hence, the goal is to show that $\int_\mathbb{R}(|x|+x^2)\nu(x,t)dx<+\infty$ for all $t>0$ given this condition is satisfied by $\nu(x,0)$. Let us begin by supposing we can apply the mild solution to $f(x)=x,x^2$. Setting $\tilde x(t)=\int_\mathbb{R}x\nu(x,t)dx$, we have
\begin{equation}
\tilde x(t)=\tilde x(0)+\int_0^t\int_\mathbb{R}\nu(x,s)m(\nu,x)x+x\sqrt{V\nu(x,s)}\dot W(x,s)dxds.
\end{equation}
Similarly, setting $\tilde{\tilde\sigma}^2(t)=\int_\mathbb{R}x^2\nu(x,t)dx$, we have
\begin{equation}
\tilde{\tilde\sigma}^2(t)=\tilde{\tilde\sigma}^2(0)+\int_0^t\int_\mathbb{R}\nu(x,s)\left(m(\nu,x)x^2+\mu\right)+x^2\sqrt{V\nu(x,s)}\dot W(x,s)dxds.
\end{equation}
Since $\bar x(t)=\tilde x(t)/N(t)$ and $\sigma^2(t)=\tilde{\tilde\sigma}^2(t)/N(t)-\bar x^2(t)$, we can use It&ocirc;'s lemma to derive SDE for $\bar x(t)$ and $\sigma^2(t)$, which we perform in SM \S\ref{SDE_DERIV}. We do not make any general assertions about the existence or uniqueness of $\bar x(t)$ or $\sigma^2(t)$.

@@

There is quite a bit we can learn from expressions (\ref{gen_eqns}). Firstly, setting $V=0$ recovers the deterministic dynamics derived in \S\ref{deterministic}. Alternatively, one can take $N(t)\to\infty$ to recover the deterministic dynamics for $\bar x(t)$ and $\sigma^2(t)$. Characteristically, we note the effect of demographic stochasticity on abundance grows with $\sqrt{N(t)}$. Hence, dividing by $N$, we find the effects of demographic stochasticity on the per-capita growth rate diminish with increased abundance. Relating the response to demographic stochasticity derived here to the effect of random genetic drift derived in classic quantitative genetic theory, if we set $\sigma^2(t)=\sigma^2$ and $N(t)=N$ constant with respect to time, then integrating the stochastic term in equation (\ref{xbar_gen}) over a single unit of time returns a normally distributed random variable with mean zero and variance equal to $V\sigma^2/N$. In particular, assuming perfect inheritance, when reproductive variance is unity ($V=1$) this random variable coincides with the effect of random genetic drift on the change in mean trait over a single generation derived using sampling arguments [@Lande1976]. There is also an interesting connection with classical population genetics. A fundamental result from early population genetic theory is the expected reduction in diversity due to the chance loss of alleles in finite populations [@Fisher1923; @Wright97]. This expected reduction in diversity due to random genetic drift is captured by the third term in the deterministic component of expression (\ref{sig2_gen}), particularly $-V\sigma^2(t)/N(t)$. The component of SDE (\ref{sig2_gen}) describing random fluctuations in $\sigma^2(t)$ is more complicated and is proportional to the root of the difference between the centralized fourth moment of $p(x,t)$ and $\sigma^4(t)$. These expressions can be used to investigate the dynamics of the mean and variance for general $\nu(x,t)$. However, in the next subsection we simplify these expressions by approximating $\nu(x,t)$ with a Gaussian curve. In SM \S\ref{SDE_DERIV} we show that under the Gaussian case $W_1,W_2$ and $W_3$ are independent.

@@

```{r, message=FALSE, echo=F, fig.align='center', fig.cap='\\label{temporal}Temporal dynamics of mean trait (top), additive genetic variance (middle) and abundance (bottom) for the scenario of weak competition (left) and strong competition (right). Red lines indicate average trend across species.'}
knitr::include_graphics('community_dynamics.png')
```

@@

## Numerical evidence of finite moments and approximate normality in the stochastic case \label{numerical}

@@

@@

We set $\tilde x(t)=\int_\mathbb{R}x\nu(x,t)$ and make use of the notation

\begin{equation}
\begin{matrix}
\|N\|_2 = \sqrt{V\int_\mathbb{R}\nu(x,t)dx}= \sqrt{VN} \\ \\
\|\tilde x\|_2 = \sqrt{V\int_\mathbb{R}x^2\nu(x,t)dx} \\ \\
\langle\tilde x,N\rangle = V\int_\mathbb{R}x\nu(x,t)dx=\bar x VN.
\end{matrix}
\end{equation}

Applying formula (\ref{SDE_deriv}) provides

@@

\begin{equation}
d\bar x= \overline{xm}dt+\frac{\mu}{2}\int_\mathbb{R}x\Delta p(x,t)dxdt+\frac{\|\tilde x\|_2}{N}d\tilde W_2-\bar x\left(\bar m dt+\sqrt\frac{V}{N}dW_1\right)-\frac{\langle\tilde x,N\rangle}{N^2}dt+\bar x\frac{\|N\|_2^2}{N^2}dt
\end{equation}

\begin{equation*}
= (\overline{xm}-\bar x\bar m)dt+\frac{\mu}{2}\int_\mathbb{R}x\Delta p(x,t)dxdt+\frac{\|\tilde x\|_2}{N}d\tilde W_2
-\bar x\sqrt{\frac{V}{N}}d W_1-V\frac{\bar x}{N}dt+V\frac{\bar x}{N}dt
\end{equation*}

\begin{equation*}
= \left(\mathrm{Cov}_t(x,m)+\frac{\mu}{2}\int_\mathbb{R}x\Delta p(x,t)dx\right)dt
+\frac{\|\tilde x\|_2}{N}d\tilde W_2-\bar x\sqrt{\frac{V}{N}}d W_1.
\end{equation*}

@@

\begin{equation}
d\bar x=\left(\mathrm{Cov}_t(x,m)+\frac{\mu}{2}\int_\mathbb{R}x\Delta p(x,t)dx\right)dt
+\sqrt{V\frac{\sigma^2}{N}}d W_2.
\end{equation}

@@

We set $\tilde{\tilde\sigma}^2(t)=\int_{\mathbb R} x^2\nu(x,t)dx$ and make use of the notation

\begin{equation}
\begin{matrix}
\|\tilde{\tilde\sigma}^2\|_2 = \sqrt{V\int_\mathbb{R}x^4\nu(x,t)dx} \\ \\
\langle\tilde{\tilde\sigma}^2,N\rangle=V\int_\mathbb{R}x^2\nu(x,t)dx=\overline{x^2}VN.
\end{matrix}
\end{equation}

Applying formula (\ref{SDE_deriv}) provides

\begin{equation}
d\tilde{\tilde\sigma}^2=\left(\overline{x^2m}N+\frac{\mu}{2}\int_\mathbb{R}x^2\Delta\nu(x,t) dx\right)dt+\|\tilde{\tilde\sigma}^2\|_2d\tilde W_3
\end{equation}

@@

\begin{multline}
d\overline{x^2}=\left(\overline{x^2m} + \frac{\mu}{2}\int_\mathbb{R}x^2\Delta p(x,t)dx\right)dt+\frac{\|\tilde{\tilde\sigma}^2\|_2}{N}d\tilde W_3
-\overline{x^2}\left(\bar m dt+\sqrt\frac{V}{N}dW_1\right)-\frac{\langle\tilde{\tilde\sigma}^2,N\rangle}{N^2}dt+\overline{x^2}\frac{\|N\|_2^2}{N^2}dt \\ \\
=\left(\overline{x^2m}-\overline{x^2}\bar mdt + \frac{\mu}{2}\int_\mathbb{R}x^2\Delta p(x,t)dx\right)dt+\frac{\|\tilde{\tilde\sigma}^2\|_2}{N}d\tilde W_3
-\overline{x^2}\sqrt{\frac{V}{N}}d W_1
-\overline{x^2}\frac{V}{N}dt+\overline{x^2}\frac{V}{N}dt \\ \\
=\left(\mathrm{Cov}_t\Big(x^2,m\Big) + \frac{\mu}{2}\int_\mathbb{R}x^2\Delta pdx \right)dt+\frac{\|\tilde{\tilde\sigma}^2\|_2}{N}d\tilde W_3-\overline{x^2}\sqrt\frac{V}{N}d W_1.
\end{multline}

@@

\begin{multline}
d\sigma^2=d\overline{x^2}-2\bar xd\bar x-(d\bar x)^2
=\left(\mathrm{Cov}_t\Big(x^2,m\Big) + \frac{\mu}{2}\int_\mathbb{R}x^2\Delta p(x,t)dx \right)dt+\frac{\|\tilde{\tilde\sigma}^2\|_2}{N}d\tilde W_3-\overline{x^2}\sqrt\frac{V}{N}d W_1 \\ \\
-2\bar x\left(\mathrm{Cov}_t(x,m)dt+\frac{\mu}{2}\int_\mathbb{R}x\Delta p(x,t)dxdt+\sqrt{\frac{V\sigma^2}{N}}d W_2\right)-\left(\mathrm{Cov}_t(x,m)dt+\frac{\mu}{2}\int_\mathbb{R}x\Delta p(x,t)dxdt+\sqrt{\frac{V\sigma^2}{N}}d W_2\right)^2 \\ \\
=\left(\mathrm{Cov}_t\Big(x^2-2\bar xx,m\Big) + \frac{\mu}{2}\int_\mathbb{R}(x^2-x\bar x)\Delta p(x,t)dx \right)dt+\frac{\|\tilde{\tilde\sigma}^2\|_2}{N}d\tilde W_3-\overline{x^2}\sqrt\frac{V}{N}d W_1-2\bar x\sqrt{\frac{V\sigma^2}{N}}d W_2-\left(\frac{V\sigma^2}{N}\right)dt \\ \\
=\left(\mathrm{Cov}_t\Big(x-\bar x)^2,m\Big) + \frac{\mu}{2}\int_\mathbb{R}(x-\bar x)^2\Delta p(x,t)dx - \frac{V\sigma^2}{N}\right)dt+\frac{\|\tilde{\tilde\sigma}^2\|_2}{N}d\tilde W_3-\overline{x^2}\sqrt\frac{V}{N}d W_1-2\bar x\sqrt{\frac{V\sigma^2}{N}}d W_2.
\end{multline}

@@

\begin{equation}
d\sigma^2=\left(\mathrm{Cov}_t\Big((x-\bar x)^2,m\Big)+\frac{\mu}{2}\int_\mathbb{R}(x-\bar x)^2\Delta p(x,t)dx-V\frac{\sigma^2}{N}\right)dt+\sqrt{V\frac{\overline{(x-\bar x)^4}-\sigma^4}{N}}d W_3.
\end{equation}

>>>>

``\int_0^t\int_\mathbb{R}f(x,s)\dot W(x,s)dxds"=\mathbf W_t(f), \ \forall \ f\in\mathscr{N}_2, \ t\geq0.

@@

To obtain a SPDE from a BBM we take a diffusion limit. There are several ways to do this, but a simple approach is to rescale the mass of individuals and time by $1/n$, diffusion by $\mu\to\mu/n$, branching rate by $\lambda\to n\lambda$, fitness by $\mathscr{W}\to\mathscr{W}^{1/n}$ and consider the limit as $n\to\infty$. Denoting the rescaled process by $X_t^{(n)}(D)$, the limiting process $\mathscr{X}_t=\lim_{n\to\infty}X_t^{(n)}$ is called a super-Brownian motion and is also a non-negative finite measure-valued process [@Watanabe1968]. However, instead of returning the number of individuals alive in a region of space, super-Brownian motion returns the _mass_ of the population concentrated in a region of space. Since we have rescaled individual mass by $1/n$ and took the limit $n\to\infty$, individuals are no longer discrete units. Instead, the particle view of the population gets replaced by a blob spread across $d$-dimensional space. In particular, the value of $\mathscr{X}_t(D)$ is a continuously varying non-negative random variable for any $t\geq0$ and $D\subset\mathbb{R}^d$.

Unfortunately, just as with cream cheese spread across too much toast, the blob perspective of the population may exhibit some complicated spatial discontinuities. However, for spatial dimension $d=1$, it turns out that $\mathscr{X}_t$ is absolutely continuous with respect to the Lebesgue measure for each $t\geq0$ [@Konno1988; @Reimers1989]. This means that we can write $\mathscr{X}_t(D)=\int_D \nu(x,t)dx$ for some density process $\nu(x,t)$. Setting $\lambda=1$ and $m=\ln\mathscr{W}$ this density process satisfies the SPDE
\begin{equation}\label{neutral_SPDE}
\frac{\partial}{\partial t}\nu(x,t)=m\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)+\sqrt{V\nu(x,t)}\dot W(x,t).
\end{equation}
Since $\nu(x,t)$ is not generally differentiable in $x$ or $t$, the derivatives in expression (\ref{neutral_SPDE}) are taken in the _weak_ sense [sensu @lawrenceevans2010]. That is, to rigorously interpret SPDE (\ref{neutral_SPDE}), we integrate the solution $\nu(x,t)$ against functions $f\in C_b^2(\mathbb{R})$, where $C_b^2(\mathbb{R})$ is the set of bounded and twice continuously differentiable functions on $\mathbb{R}$. Hence, equation (\ref{neutral_SPDE}) is just an abbreviation for

\begin{multline}\label{neutral_int}
\int_\mathbb{R}\nu(x,t)f(x)dx-\int_\mathbb{R}\nu(x,0)f(x)dx=\int_0^t\int_\mathbb{R}\nu(x,s)\left(mf(x)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}f(x)\right)dsdx \\
+\int_0^t\int_\mathbb{R}f(x)\sqrt{V\nu(x,s)}\dot W(x,s)dxds, \ \forall \ f\in C_b^2(\mathbb{R}).
\end{multline}
This expression is referred to as the _mild_ solution of (\ref{neutral_SPDE}). For more on the general theory of SPDE see Walsh [-@Walsh] and Da Prato and Zabczyk [-@DaPrato2014]. Note that since $\nu(x,t)$ is the density of a finite measure, it is integrable for each $t\geq0$. Thus, since for some $M>0$, $|f(x)|\leq M$ for every $x\in\mathbb{R}$, setting $\varphi(x,t)=f(x)\sqrt{V\nu(x,t)}$ implies $\varphi\in\mathscr{N}_2$. Hence, the white noise integral on the right-hand side of equation (\ref{neutral_int}) can be understood using the heuristics introduced above. Evaluating equation (\ref{neutral_int}) in the particular case of $f(x)\equiv1$ returns the total mass process, which we refer to as the total abundance $N(t)$.

@@

where
\begin{equation}
\bar m(t)=\frac{1}{N(t)}\int_\mathbb{R}m(\nu,x)\nu(x,t)dx,
\end{equation}
and
\begin{equation}
\int_0^td\hat{\mathbf{W}}_s(\sqrt{\nu(x,s)})=\int_0^t\int_\mathbb{R}\frac{\sqrt{\nu(x,s)}}{\sqrt{\int_\mathbb{R}\nu(x,s)dx}}\dot W(x,s)dxds.
\end{equation}
Setting $W_1(t)=\hat{\mathbf{W}}_t(\sqrt{\nu(x,t)})$, we can use traditional stochastic differential notation to write
\begin{equation}
dN=\bar mNdt+\sqrt{VN}dW_1.
\end{equation}
To find the associated SDE for $\bar x(t)$ and $\sigma^2(t)$, we want to repeat the same approach for $f(x)=x,x^2$ and apply It&ocirc;'s lemma. However, for these cases $f\notin C^2_b(\mathbb{R})$ since $f$ will not be bounded. However, if we can show that $\int_\mathbb{R}(|x|+x^2+x^4)\nu(x,t)dx<+\infty$ for all $t>0$ given this condition is satisfied by $\nu(x,0)$, then we can apply the mild solution of (\ref{SPDE}) to derive SDE for $\bar x(t)$ and $\sigma^2(t)$. To illustrate, let us suppose this is the case. Setting $\tilde x(t)=\int_\mathbb{R}x\nu(x,t)dx$, we have
\begin{equation}
\tilde x(t)=\tilde x(0)+\int_0^t\int_\mathbb{R}\nu(x,s)m(\nu,x)x+x\sqrt{V\nu(x,s)}\dot W(x,s)dxds.
\end{equation}
Similarly, setting $\tilde{\tilde\sigma}^2(t)=\int_\mathbb{R}x^2\nu(x,t)dx$, we have
\begin{equation}
\tilde{\tilde\sigma}^2(t)=\tilde{\tilde\sigma}^2(0)+\int_0^t\int_\mathbb{R}\nu(x,s)\left(m(\nu,x)x^2+\mu\right)+x^2\sqrt{V\nu(x,s)}\dot W(x,s)dxds.
\end{equation}
Since $\bar x(t)=\tilde x(t)/N(t)$ and $\sigma^2(t)=\tilde{\tilde\sigma}^2(t)/N(t)-\bar x^2(t)$, we can use It&ocirc;'s lemma to derive SDE for $\bar x(t)$ and $\sigma^2(t)$, which we perform in SM \S\ref{SDE_DERIV}. We do not make any general assertions about the existence or uniqueness of $\bar x(t)$ or $\sigma^2(t)$.

@@

There is quite a bit we can learn from expressions (\ref{gen_eqns}). Firstly, setting $V=0$ recovers the deterministic dynamics derived in \S\ref{deterministic}. Alternatively, one can take $N(t)\to\infty$ to recover the deterministic dynamics for $\bar x(t)$ and $\sigma^2(t)$. Characteristically, we note the effect of demographic stochasticity on abundance grows with $\sqrt{N(t)}$. Hence, dividing by $N$, we find the effects of demographic stochasticity on the per-capita growth rate diminish with increased abundance. Relating the response to demographic stochasticity derived here to the effect of random genetic drift derived in classic quantitative genetic theory, if we set $\sigma^2(t)=\sigma^2$ and $N(t)=N$ constant with respect to time, then integrating the stochastic term in equation (\ref{xbar_gen}) over a single unit of time returns a normally distributed random variable with mean zero and variance equal to $V\sigma^2/N$. In particular, assuming perfect inheritance, when reproductive variance is unity ($V=1$) this random variable coincides with the effect of random genetic drift on the change in mean trait over a single generation derived using sampling arguments [@Lande1976]. There is also an interesting connection with classical population genetics. A fundamental result from early population genetic theory is the expected reduction in diversity due to the chance loss of alleles in finite populations [@Fisher1923; @Wright97]. This expected reduction in diversity due to random genetic drift is captured by the third term in the deterministic component of expression (\ref{sig2_gen}), particularly $-V\sigma^2(t)/N(t)$. The component of SDE (\ref{sig2_gen}) describing random fluctuations in $\sigma^2(t)$ is more complicated and is proportional to the root of the difference between the centralized fourth moment of $p(x,t)$ and $\sigma^4(t)$.

These expressions can be used to investigate the dynamics of the mean and variance for general $\nu(x,t)$. However, in the next subsection we simplify these expressions by approximating $\nu(x,t)$ with a Gaussian curve. By assuming $\nu(x,t)$ is Gaussian for $t\geq0$, we guarantee the existence of $\bar x(t)$ and $\sigma^2(t)$ for all $t\geq0$. Furthermore, in SM \S\ref{SDE_DERIV} we show that under the Gaussian case $W_1,W_2$ and $W_3$ are independent.

@@

```{r, message=FALSE, echo=F, fig.align='center', fig.cap='\\label{temporal}Temporal dynamics of mean trait (top), additive genetic variance (middle) and abundance (bottom) for the scenario of weak competition (left) and strong competition (right). Red lines indicate average trend across species.'}
knitr::include_graphics('community_dynamics.jpg')
```

@@

## Simulating the rescaled process and numerical evidence of approximate normality \label{numerical}

@@

Picking up from \S\ref{stochastic} of the main text, we have
\begin{equation}
\tilde x(t)=\int_\mathbb{R}x\nu(x,t)dx, \ \ \tilde{\tilde\sigma}^2(t)=\int_\mathbb{R}x^2\nu(x,t)dx
\end{equation}
and
\begin{equation}\label{xtilde}
\tilde x(t)=\tilde x(0)+\int_0^t\int_\mathbb{R}\nu(x,s)m(\nu,x)x+x\sqrt{V\nu(x,s)}\dot W(x,s)dxds,
\end{equation}
\begin{equation}\label{stilde}
\tilde{\tilde\sigma}^2(t)=\tilde{\tilde\sigma}^2(0)+\int_0^t\int_\mathbb{R}\nu(x,s)\left(m(\nu,x)x^2+\mu\right)+x^2\sqrt{V\nu(x,s)}\dot W(x,s)dxds.
\end{equation}

@@

We make use of the notation

\begin{equation}
\left\{\begin{matrix}
\|N\|_2 = \sqrt{V\int_\mathbb{R}\nu(x,t)dx}= \sqrt{VN} \\ \\
\|\tilde x\|_2 = \sqrt{V\int_\mathbb{R}x^2\nu(x,t)dx} \\ \\
\langle\tilde x,N\rangle = V\int_\mathbb{R}x\nu(x,t)dx=\bar x VN.
\end{matrix}\right.
\end{equation}

Rewriting formula (\ref{xtilde}) as an SDE provides

@@

\begin{equation}
d\bar x= \overline{xm}dt+\frac{\|\tilde x\|_2}{N}d\tilde W_2-\bar x\left(\bar m dt+\sqrt\frac{V}{N}dW_1\right)-\frac{\langle\tilde x,N\rangle}{N^2}dt+\bar x\frac{\|N\|_2^2}{N^2}dt
\end{equation}

\begin{equation*}
= (\overline{xm}-\bar x\bar m)dt+\frac{\|\tilde x\|_2}{N}d\tilde W_2
-\bar x\sqrt{\frac{V}{N}}d W_1-V\frac{\bar x}{N}dt+V\frac{\bar x}{N}dt
\end{equation*}

\begin{equation*}
=\mathrm{Cov}_t(x,m)
+\frac{\|\tilde x\|_2}{N}d\tilde W_2-\bar x\sqrt{\frac{V}{N}}d W_1.
\end{equation*}

@@

\begin{equation}
d\bar x=\mathrm{Cov}_t(x,m)dt
+\sqrt{V\frac{\sigma^2}{N}}d W_2.
\end{equation}

@@

We make use of the notation

\begin{equation}
\left\{\begin{matrix}
\|\tilde{\tilde\sigma}^2\|_2 = \sqrt{V\int_\mathbb{R}x^4\nu(x,t)dx} \\ \\
\langle\tilde{\tilde\sigma}^2,N\rangle=V\int_\mathbb{R}x^2\nu(x,t)dx=\overline{x^2}VN.
\end{matrix}\right.
\end{equation}

Applying formula (\ref{stilde}) provides

\begin{equation}
d\tilde{\tilde\sigma}^2=\left(\overline{x^2m}N+\mu N\right)dt+\|\tilde{\tilde\sigma}^2\|_2d\tilde W_3
\end{equation}

@@

\begin{multline}
d\overline{x^2}=\left(\overline{x^2m} + \mu \right)dt+\frac{\|\tilde{\tilde\sigma}^2\|_2}{N}d\tilde W_3
-\overline{x^2}\left(\bar m dt+\sqrt\frac{V}{N}dW_1\right)-\frac{\langle\tilde{\tilde\sigma}^2,N\rangle}{N^2}dt+\overline{x^2}\frac{\|N\|_2^2}{N^2}dt \\ \\
=\left(\overline{x^2m}-\overline{x^2}\bar mdt + \mu\right)dt+\frac{\|\tilde{\tilde\sigma}^2\|_2}{N}d\tilde W_3
-\overline{x^2}\sqrt{\frac{V}{N}}d W_1
-\overline{x^2}\frac{V}{N}dt+\overline{x^2}\frac{V}{N}dt \\ \\
=\left(\mathrm{Cov}_t\Big(x^2,m\Big) + \mu \right)dt+\frac{\|\tilde{\tilde\sigma}^2\|_2}{N}d\tilde W_3-\overline{x^2}\sqrt\frac{V}{N}d W_1.
\end{multline}

@@

\begin{multline}
d\sigma^2=d\overline{x^2}-2\bar xd\bar x-(d\bar x)^2
=\left(\mathrm{Cov}_t\Big(x^2,m\Big) + \mu \right)dt+\frac{\|\tilde{\tilde\sigma}^2\|_2}{N}d\tilde W_3-\overline{x^2}\sqrt\frac{V}{N}d W_1 \\ \\
-2\bar x\left(\mathrm{Cov}_t(x,m)+\mu dt+\sqrt{\frac{V\sigma^2}{N}}d W_2\right)-\left(\mathrm{Cov}_t(x,m)dt+\mu dt+\sqrt{\frac{V\sigma^2}{N}}d W_2\right)^2 \\ \\
=\left(\mathrm{Cov}_t\Big(x^2-2\bar xx,m\Big) + \frac{\mu}{2}\int_\mathbb{R}(x^2-x\bar x)\Delta p(x,t)dx \right)dt+\frac{\|\tilde{\tilde\sigma}^2\|_2}{N}d\tilde W_3-\overline{x^2}\sqrt\frac{V}{N}d W_1-2\bar x\sqrt{\frac{V\sigma^2}{N}}d W_2-\left(\frac{V\sigma^2}{N}\right)dt \\ \\
=\left(\mathrm{Cov}_t\Big(x-\bar x)^2,m\Big) + \frac{\mu}{2}\int_\mathbb{R}(x-\bar x)^2\Delta p(x,t)dx - \frac{V\sigma^2}{N}\right)dt+\frac{\|\tilde{\tilde\sigma}^2\|_2}{N}d\tilde W_3-\overline{x^2}\sqrt\frac{V}{N}d W_1-2\bar x\sqrt{\frac{V\sigma^2}{N}}d W_2.
\end{multline}

@@

\begin{equation}
d\sigma^2=\left(\mathrm{Cov}_t\Big((x-\bar x)^2,m\Big)+\mu-V\frac{\sigma^2}{N}\right)dt+\sqrt{V\frac{\overline{(x-\bar x)^4}-\sigma^4}{N}}d W_3.
\end{equation}
